---
title: "Sharing Data with {pins}"
subtitle: "Versioned Data Collaboration in R"
author: "Joe Corra, OAR | OCAP | CCPD | CPB"
date: today
format:
  revealjs:
    theme: moon
    slide-number: true
    chalkboard: true
    preview-links: auto
---

## The Problem

How do you share data with your team?

. . .

Common pain points:

- Email attachments
- Shared drives with messy file names
- `final_final_v3.csv`
- "Which version is current?"
- Lost context and metadata

## Enter {pins}

A package for **publishing and sharing** data, models, and other R objects

. . .

Key benefits:

- **Versioned**: automatic version tracking
- **Discoverable**: metadata and search
- **Reproducible**: programmatic access
- **Collaborative**: share across teams

## Installation

```{r}
#| eval: false
#| echo: true

# Install from CRAN
install.packages("pins")

# Load the package
library(pins)
```

# Basic Concepts

## What is a Pin?

A **pin** is a stored R object with metadata

. . .

You can pin:

- Data frames
- Trained models
- Lists
- Spatial objects (sf)
- Any R object

- Choose the file type ("csv", "json", "rds", "parquet", "arrow", "qs")

## What is a Board?

A **board** is where pins live

. . .

Think of it as a storage location:

- Local folder
- Network drive
- RStudio Connect
- Cloud storage (S3, Azure, GCS)
- URL (read-only)

## Three Main Functions

```{r}
#| eval: false
#| echo: true

# Write a pin
board %>% pin_write(data, "my-dataset")

# Read a pin
data <- board %>% pin_read("my-dataset")

# List available pins
board %>% pin_list()
```

# Live Demo {background-color="#40666e"}

## Create a Local Board

```{r}
#| eval: false
#| echo: true

library(pins)

# Create a board in a local folder
board <- board_folder("~/my-pins")
```

## Pin Some Data

```{r}
#| eval: false
#| echo: true

# Create sample data
sales_data <- data.frame(
  month = month.name,
  sales = sample(1000:5000, 12),
  region = sample(c("North", "South", "East", "West"), 12, replace = TRUE)
)

# Pin it with metadata
board %>% pin_write(
  sales_data,
  name = "monthly-sales",
  title = "Monthly Sales Data",
  description = "Sales by month and region for 2024"
)
```

## Read the Pin

```{r}
#| eval: false
#| echo: true

# Read the pin
retrieved_data <- board %>% pin_read("monthly-sales")

# View it
head(retrieved_data)
```

## Update and Version

```{r}
#| eval: false
#| echo: true

# Update the data
sales_data$sales <- sales_data$sales * 1.1

# Write again (creates new version)
board %>% pin_write(sales_data, "monthly-sales")

# Check versions
board %>% pin_versions("monthly-sales")
```

## Access Specific Versions

```{r}
#| eval: false
#| echo: true

# Read a specific version
old_data <- board %>% pin_read(
  "monthly-sales",
  version = "20240115T103000Z"  # example timestamp
)

# Prune old versions (keep last 5)
board %>% pin_versions_prune("monthly-sales", n = 5)
```

## Metadata and Discovery

```{r}
#| eval: false
#| echo: true

# View metadata
board %>% pin_meta("monthly-sales")

# Search for pins
board %>% pin_search("sales")

# Delete a pin
board %>% pin_delete("monthly-sales")
```

# Real-World Use Cases

## Use Case 1: Reference Data

Share lookup tables and reference datasets:

- Fiscal calendars
- Product codes
- Geographic mappings
- Standard assumptions

. . .

**Benefit**: Everyone uses the same version, easily updated centrally

## Use Case 2: Analysis-Ready Data

One person cleans messy data, others analyze:

```{r}
#| eval: false
#| echo: true

# Data engineer cleans and pins
clean_data <- raw_data %>% 
  clean_names() %>%
  filter(!is.na(key_field))

board %>% pin_write(clean_data, "analysis-ready-sales")

# Analysts read and use
analysis_data <- board %>% pin_read("analysis-ready-sales")
```

## Use Case 3: Model Deployment

Store trained models for production use:

```{r}
#| eval: false
#| echo: true

# Train and pin a model
model <- lm(sales ~ region + month, data = sales_data)
board %>% pin_write(model, "sales-forecast-model")

# Load in production script
model <- board %>% pin_read("sales-forecast-model")
new_predictions <- predict(model, new_data)
```

## Use Case 4: Caching Expensive Operations

Cache API calls or database queries:

```{r}
#| eval: false
#| echo: true

# Check if cached version exists
if (!"api-data" %in% board %>% pin_list()) {
  # Expensive API call
  api_data <- fetch_from_api()
  board %>% pin_write(api_data, "api-data")
} else {
  # Use cached version
  api_data <- board %>% pin_read("api-data")
}
```

# When NOT to Use Pins

## Limitations to Consider

Pins may not be the best choice for:

. . .

- **Very large datasets** (multi-GB) - consider databases or direct Arrow files
- **High-frequency updates** - real-time data needs a database
- **Complex data pipelines** - {targets} might be better
- **Formal data warehousing** - enterprise data infrastructure
- **Extremely sensitive data** - may need database with audit logs

## When Pins Shines

The "sweet spot":

- **Medium-sized datasets** (MBs to low GBs)
- **Sharing between R users**
- **Objects that don't fit in databases** (models, lists, sf objects)
- **Quick collaboration** without infrastructure overhead

. . .

**Think of it as**: Too complex for email, too simple for a full database

# Beyond Local Boards

## Board Types

Different boards for different needs:

```{r}
#| eval: false
#| echo: true

# Network drive
board <- board_folder("//server/shared/team-pins")

# RStudio Connect
board <- board_connect()

# AWS S3
board <- board_s3("my-bucket")

# Azure
board <- board_azure("container", "account")

# Google Cloud
board <- board_gcs("bucket")
```

## board_url() for Read-Only Sharing

Publish pins via simple web hosting:

```{r}
#| eval: false
#| echo: true

# Connect to a board hosted at a URL
board <- board_url(c(
  "monthly_sales" = "https://example.com/data/sales.rds",
  "customer_ref" = "https://example.com/data/customers.rds"
))

# Read a pin
sales_data <- board %>% pin_read("monthly_sales")
```

. . .

Great for: GitHub Pages, internal web servers, publishing reference data

## Recommendations for Your Team

**Start simple**: `board_folder()` on a shared network drive

. . .

**If you have RStudio Connect**: Use it! Purpose-built for this.

. . .

**Cloud-enabled**: S3/Azure/GCS boards scale well

. . .

**The best board depends on what your IT already supports**

# Common Questions

## "Why not just use GitHub?"

GitHub works great for small CSVs in code repos!

. . .

Pins adds value for:

- Non-CSV objects (models, lists, sf data)
- Separating data updates from code commits
- No git overhead for data consumers
- Automatic metadata and versioning
- Easy programmatic access

## "Why not SharePoint?"

SharePoint absolutely works!

. . .

But consider:

- **Fragile file paths** that break when folders reorganize
- **Manual version management** (file naming or web interface)
- **No R-native integration** (just reading from paths)
- **Discovery** requires browsing folders

. . .

**Pins solution**: `pin_read("clean-sales")` works the same for everyone

## "Can I download any CSV from a website?"

Not really the purpose of `board_url()`

. . .

For raw CSVs on websites, just use:

```{r}
#| eval: false
#| echo: true

data <- read.csv("https://example.com/data/file.csv")
```

. . .

`board_url()` is for datasets **specifically published as pins**

# Working with Arrow/Parquet

## Parquet, Feather, Arrow - Brief Intro

**Parquet**: Columnar file format, compressed, great for storage

**Feather**: Faster, less compressed, good for temporary use

**Arrow**: The ecosystem/engine that works with both formats

. . .

All work across R, Python, SQL, Spark, etc.

## Pins + Parquet

```{r}
#| eval: false
#| echo: true

library(arrow)

# Write as parquet
board %>% pin_write(my_data, "dataset", type = "parquet")

# Read (still loads into memory)
data <- board %>% pin_read("dataset")
```

. . .

**For larger data**, use `pin_download()`:

```{r}
#| eval: false
#| echo: true

# Download without loading
path <- board %>% pin_download("dataset")

# Use Arrow for lazy evaluation
arrow::open_dataset(path) %>%
  filter(year == 2024) %>%
  collect()
```

# Wrap Up

## Key Takeaways

1. Pins solves version control and discovery for data sharing
2. Simple API: `pin_write()`, `pin_read()`, `pin_list()`
3. Multiple board types fit different infrastructure
4. Best for medium-sized data and R-to-R collaboration
5. Complements (not replaces) Git, SharePoint, databases

## Resources

- Package documentation: [pins.rstudio.com](https://pins.rstudio.com)
- GitHub: [github.com/rstudio/pins-r](https://github.com/rstudio/pins-r)
- RStudio Community: [community.rstudio.com](https://community.rstudio.com)

## Questions?

Thank you!

---

**Contact**: your.email@company.com

**Slides**: [link to slides if hosted]
